From Parametricity to Conservation Laws, via Noether's Theorem

### 2013-07-10 mechanics/paper.tex
Work on the introduction

@@

\begin{enumerate}
\item Symmetry in programming and physics
  \begin{enumerate}
  \item In programming: replacement of one data type by another
  \item In physics: invariance under change of unobservable
    properties. FIXME: check this.
  \item What unifies these, at a conceptual level? Inability to
    distinguish between situations, given the tools provided. In both
    cases, this leads to interesting properties of such systems. In
    programming, we get ``free theorems''. In Physics, via Noether's
    theorem, we get ``conservation laws''.
  \item In this paper, we show that we can define a programming
    language for classical mechanics, based on System F$\omega$, where
    the abstraction theorem directly fulfils the hypothesis of
    Noether's theorem, allowing us to derive conservation laws.
  \end{enumerate}
\item Parametricity: polymorphic programs have symmetry properties
  \begin{enumerate}
  \item Example of a free theorem
  \item Example of a geometric free theorem (the area of a triangle
    example)
  \end{enumerate}
\item Noether's theorem: symmetry properties of descriptions of
  systems lead to conservation laws
\item Let's put the two together
  \begin{enumerate}
  \item Need invariance properties of smooth functions
  \item So we extend System F$\omega$ with 
  \end{enumerate}
\item We present the classical mechanics and Noether's theorem first,
  since it is likely to be more unfamiliar to readers.
\end{enumerate}

We restate the results of Atkey, Johann and Kennedy in the more
general setting of System F$\omega$... which is good because it allows
for a deeper interpretation of what the unit kinds are. Also, we can
introduce useful stuff like nat-indexed types without too much fuss.

\paragraph{Summary of the Paper}

@@

\paragraph{Invariance of the Action} Let
\begin{displaymath}
  \mathcal{S}[q;a;b] = \int_a^b L(t,q(t),\dot{q}(t)) \mathit{dt}
\end{displaymath}
be the action of some physical system described by the Lagrangian
$L$. Assume a differentiable invertible function $\tau : \mathbb{R}
\to \mathbb{R}$ that transforms time in some way and a function
$\sigma : \mathbb{R} \times \mathbb{R}^n \to \mathbb{R}^n$ that
transforms the vector of generalised coordinates $\vec{q}$ at time $t$
to another vector of generalised coordinates $\sigma(t,\vec{q})$.

We say that the action $\mathcal{S}$ is invariant under the
transformations $\tau$ and $\sigma$ if it is the case that for all
$a$, $b$ and $q$ that:
\begin{displaymath}
  \int_a^b L(t,q,\dot{q}) \mathit{dt} = \int_{\tau(a)}^{\tau(b)} L(t', \sigma(\tau^{-1}(t'), q(t')), \mathbf{FIXME}) \mathit{dt'}
\end{displaymath}
FIXME: Remark on the use of $t$ and $t'$ in the integrals. FIXME:
state that all we are doing is transforming the time and the path by
the transformations $\tau$ and $\sigma$ respectively. State why it
will be useful for the

Remark on how to give a simpler version of the invariance principle by
applying some simple transformations of integrals (see notes).

@@

The interpretation of kinds as reflexive graphs, and small kinds as
small reflexive graphs is a key property of our semantics that we will
maintain as we add additional kinds in Sections XXX below.

@@

In Section~\ref{sec:refl-graphs-for-fomega}, we only had a single base
kind: the kind $*$ of proper types.

Natural numbers as a discrete kind. The type of length-indexed
vectors.

@@

@@

\subsection{Group(oid) Kinds}
\label{sec:groupoid-kinds}

@@

\begin{enumerate}
\item Each (matrix) group gives a kind, which we can use as parameter
  to a type.
\item The collection of cartesian spaces, with diffeomorphisms as the
  edges.
\item Do units of measure as an example (include one of the results
  from that paper, e.g., type iso)
\item Mention groupoid kinds, and link to Hofmann and Streicher. Do
  finite sets and bijections as an example.
\end{enumerate}

\paragraph{Why not Groupoids Everywhere?} Explain why we aren't using
groupoids as kinds: don't need to, doesn't directly generalise
relational parametricity, see Robinson's paper. Distance-indexed
types.

@@

\subsection{Extending System F$\omega$ for Classical Mechanics}

@@

\begin{enumerate}
\item Kind of cartesian spaces, parameterised by a matrix group and a
  translation group
\item Smooth functions between spaces: motivate this by looking at the
  requirements of the Euler-Lagrange equations.
\item Primitives for dealing with the geometry of the spaces (inner
  product, vector space operations, cross product(?), trigonometric
  functions (at least sin and cos)).\footnote{FIXME: fix the notation of $\forall$}
  \begin{displaymath}
    \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
      \sin &:& \forall z \mathord: \mathsf{Z}.~\mathbb{R}\langle 2\pi \cdot z, 1 \rangle \smoothrightarrow \mathbb{R} \langle 0, 1 \rangle \\
      \cos &:& \forall z \mathord: \mathsf{Z}.~\mathbb{R}\langle 2\pi \cdot z, 1 \rangle \smoothrightarrow \mathbb{R} \langle 0, 1 \rangle \\
    \end{array}
  \end{displaymath}
\end{enumerate}

\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \mathrm{exp} &:& \forall t \mathord: \mathsf{T}(1).~\mathbb{R}\langle t, 1 \rangle \smoothrightarrow \mathbb{R}\langle 0, \mathit{exp}~t \rangle
  \end{array}
\end{displaymath}

Group homomorphisms:
\begin{itemize}
\item $\mathit{exp} : \mathsf{T}(1) \to \mathsf{GL}(1)$
\item $\mathit{scale}_n : \mathsf{GL}(1) \to \mathsf{GL}(n)$
\item $\mathit{ortho}_n : \mathsf{O}(n) \to \mathsf{GL}(n)$
\item $\underline{c}\cdot- : \mathsf{Z} \to \mathsf{T}(1)$
\end{itemize}
Group operations on translations (for all $n$), and on
$\mathsf{GL}(1)$ (no abs, because it is not smooth).

\begin{displaymath}
  \kappa ::= \cdots \sepbar \mathsf{CartSp} \sepbar \mathsf{T}(n) \sepbar \mathsf{GL}(n) \sepbar \mathsf{O}(n) \sepbar \mathsf{Z} \sepbar XXX
\end{displaymath}

\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    (\smoothrightarrow) &:& \mathsf{CartSp} \to \mathsf{CartSp} \to * \\
    \mathbb{R}^n &:& \mathsf{T}(n) \to \mathsf{GL}(n) \to \mathsf{CartSp} \\
    (\times) &:& \mathsf{CartSp} \to \mathsf{CartSp} \to \mathsf{CartSp}
  \end{array}
\end{displaymath}

@@

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle \mathit{scale}_n(s)\mathit{ortho}_n(o), 0 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle \mathit{scale}_n(s)\mathit{ortho}_n(o), 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 \cdot e_2 : \mathbb{R}\langle\mathit{scale}_n(s)^2, 0 \rangle}
  \end{mathpar}
  \textbf{Transcendental functions}
  \begin{mathpar}
    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 1, t \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{exp}~e : \mathbb{R}\langle \mathit{exp}~t, 0 \rangle}

@@

\newpage

Overall plan:
\begin{enumerate}
\item A type system for writing smooth functions between smooth
  manifolds, subject to invariance with respect to certain Lie groups
\item Use this to write Lagrangians that describe classical mechanics
  situations, in such a way that Noether's theorem can be applied.
\item Need to explain:
  \begin{enumerate}
  \item Lagrangians
  \item Euler-Lagrange equations for $L(t,q,q')$
    \begin{displaymath}
      \frac{\mathrm{d}}{\mathrm{d}t} \frac{\partial L}{\partial q'} - \frac{\partial L}{\partial q} = 0
    \end{displaymath}
  \item Lie groups (manifolds!?), and their infinitesimal generators
    (see the exposition in Olver's book).
  \item Noether's theorem, with (lots of) examples. Explain what was
    in Noether's original paper (first theorem vs. second theorem).
  \end{enumerate}
  Interleave the presentation of the maths/physics and the type
  system?
\item Stick to being relatively uncomplicated with respect to the
  manifold stuff? Or use dependent types to handle the tangent spaces
  ``properly''?
\item Operations we want (in order to do the examples): distance
  between points (arclength/geodesic length), trigonometric functions
  (what continuous invariance properties do these have?),
\item Possibly use a variant of System F$\omega$ as the
  superstructure: System F$\omega$ plus a sub-language for dealing
  with smooth functions (which, in principle, could be handled just by
  using combinators to build something in a arrow-style). System
  F$\omega$ gives us a lot of power, and means that we don't need to
  go anywhere near dependent types...
  \begin{enumerate}
  \item So manifolds are a kind
  \item And so are Lie groups (?), not really, they have a different
    status (due to the identity extension requirement)
  \item Flat spaces are type operators from groups that act on them to
    manifolds
  \item Have a type constructor that builds the space of infinitely
    differentiable functions between two spaces
  \item Elaboration process that desugars a nicer syntax for smooth
    functions...
  \item Possibly a bit of natural numbers and vectors to allow for
    $n$-body problems...
  \item Need to restrict to predicativity in order to get classical
    mathematics? (probably should just forget about type
    quantification, and restrict the set of things we can quantify
    over to certain ``small'' kinds; just the groups?). Mention that
    going for SDG as a basis might offer a way out of this.
  \item Sub-kinding?
  \end{enumerate}
\item Examples:
  \begin{enumerate}
  \item $n$-body problem in $\mathbb{R}^3$, with all the consequent
    conservation laws.
  \item Central force problem
  \item Pendulums
  \item Rigid body problem? Use quotients in the definitions of the
    trigonometric functions? $\sin : \forall n :
    \mathbb{Z}.~\mathbb{R}\langle2\pi n\rangle \to \mathbb{R}\langle 1
    \rangle$. This gives us a way of doing circles in a neat way.
  \end{enumerate}
\end{enumerate}

\begin{enumerate}
\item Be clear that we don't really expect these programs to
  necessarily be (directly) executable.
\item Manifolds?
  \begin{enumerate}
  \item Spheres
  \item Tangent spaces
  \item 'Flat' spaces
  \item (Quotient spaces?) (have a look at Olver's book)
  \end{enumerate}
\end{enumerate}

Groups and their operations:
\begin{enumerate}
\item General linear group
\item Orthogonal group
\item Rotations around certain axes (in 
\end{enumerate}

Types (maybe include dependent types, for $n$-ary sized problems?)
\begin{displaymath}
  \begin{array}{rcl}
    M,M' & ::= & \mathbb{R}^n\langle e \rangle \sepbar \Sigma x : M.~M' \sepbar T_xM \\
    A, B & ::= & S \sepbar C^\infty(S,T) \sepbar A \times B \sepbar \forall G.~A \sepbar A \to B \sepbar A + B
  \end{array}
\end{displaymath}

\begin{mathpar}
  \inferrule*
  {\Gamma \sepbar x : S \vdash M :^\infty T}
  {\Gamma \vdash \lambda^\infty x : S.~M : C^\infty(S,T)}
\end{mathpar}

So a Lagrangian is something of type:
\begin{displaymath}
  \forall G:\mathsf{E}(3).~C^\infty(\mathbb{R} \times \mathbb{R}^3\langle G \rangle \times \mathbb{R}^3\langle G \rangle, \mathbb{R})
\end{displaymath}

\section{Lagrangian Mechanics}

\section{Symmetry and Conservation Laws}

\section{A Type System for Lagrangians}

Syntax and Semantics, and the abstraction theorem...

\section{Field Theory}

>>>>

\category{CR-number}{subcategory}{third-level}

\terms
term1, term2

\keywords
keyword1, keyword2

@@

Reynolds' theory of relational parametricity \cite{reynolds82} tells
us that parametrically polymorphic programs automatically satisfy
invariance properties. Such invariance properties are often called
\emph{Free Theorems}, after Wadler \cite{wadler89theorems}, since they
follow ``for free'' from the types of programs, rather than through
detailed study of the program text itself. A classic example,
presented by Wadler, is the free theorem for programs $f$ with the
following type:
\begin{displaymath}
  f : \forall \alpha.~\mathsf{List}~\alpha \to \mathsf{List}~\alpha
\end{displaymath}
Such programs takes lists of $\alpha$s, for any type $\alpha$, to
lists of $\alpha$s. Using Reynolds' theory of relational
parametricity, Wadler showed that \emph{any} $f$ with this type
satisfies the following property:
\begin{displaymath}
  \forall \alpha, \beta, g : \alpha \to \beta, l : \mathsf{List}~\alpha.~\mathrm{map}~g~(f~[\alpha]~l) = f~[\beta]~(\mathrm{map}~g~l)
\end{displaymath}
Thus, any $f$ with the type given above commutes with mapping some
arbitrary function $g$ over lists. Thinking in terms of abstract data
types and change of data representation, this free theorem states that
$f$ is invariant under change of data representation from an arbitrary
type $\alpha$ to another arbitrary type $\beta$, via $g$.

Using invariance under change to derive useful properties is a concept
much older than programming languages. In physics, Noether's theorem
\cite{noether} provides a general way to derive conserved properties
of physical systems from

FIXME: example of translation invariance, with the coupled spring example
\begin{equation}\label{eq:coupled-spring-langrangian}
  L(t,q_1,q_2,\dot{q_1},\dot{q_2}) = \frac{1}{2}m(\dot{q_1}^2 + \dot{q_2}^2) - \frac{1}{2}k(q_1 - q_2)^2
\end{equation}
As a general rule, Lagrangians for classical mechanical systems take
the form $L = T - V$, where $T$ is the kinetic energy of the system
and $V$ is the potential energy. From the Lagrangian, using the
principle of stationary action, we can derive the following two
equations of motion for this system. (We describe the principle of
stationary action, and the process for deriving the equations of
motion fully in Section~\ref{sec:conservation-laws-from-symmetry}.)
For this system, the equations of motion are a pair of ordinary
differential equations (ODEs) that describe how the positions and
velocities of the particles evolve over time:
\begin{displaymath}
  FIXME
\end{displaymath}
We could now proceed to solve these ODEs to further analyse the
behaviour of this system. However, Noether's theorem gives us a
powerful way of gaining insight into properties of these ODEs ``for
free'', without necessarily having to find solutions to them.

The Lagrangian (\ref{eq:coupled-spring-langrangian}) does not refer to
any fixed point in space; only the relative distance between the two
particles, along with their velocities, is relevant. Therefore, the
Langrangian (\ref{eq:coupled-spring-langrangian}) is invariant under
translation in space by some arbitrary displacement $x$:
\begin{equation}\label{eq:coupled-spring-translation-invariance}
  L(t,q_1,q_2,\dot{q_1},\dot{q_2}) = L(t,q_1+x,q_2+x,\dot{q_1},\dot{q_2})
\end{equation}
By Noether's theorem, invariance under spatial translation implies
that the linear momentum of the whole system is constant for all
time. For this system, conservation of linear momentum is stated
mathematically like so:
\begin{equation}\label{eq:coupled-spring-linear-momentum}
  FIXME
\end{equation}
In general, Noether's theorem gives us a way of deriving conserved
properties like (\ref{eq:coupled-spring-linear-momentum}) from
invariance properties like
(\ref{eq:coupled-spring-translation-invariance}). In this case, we
have used invariance under translation in space to derive conservation
of linear momentum. Other common examples include invariance under
translation in time, yielding conservation of energy, and invariance
under rotation, yield conservation of angular momentum. We will see
examples of each of these kinds of invariance, and consequent
conservation laws, in Section~\ref{sec:examples}.

The invariance property stated in
Equation~(\ref{eq:coupled-spring-translation-invariance}) is highly
reminisicent of the

In this paper, we present a way of using a generalised version of
Reynolds' theory of relational parametricity to prove the geometric
invariance properties required for the hypotheses of Noether's
theorem. We build an extension of System F$\omega$ suitable for
writing invariant Lagrangians. In our system, the Lagrangian
(\ref{eq:coupled-spring-langrangian}) we gave above for describing a
system of two particles coupled by a spring will have the type:
\begin{displaymath}
  \forall x \mathord: \mathsf{T}(1).~C^\infty(\mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, x \rangle \times \mathbb{R}\langle 1, x \rangle \times \mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, 0 \rangle, \mathbb{R}\langle 1, 0 \rangle)
\end{displaymath}
where... FIXME

Atkey, Johann and Kennedy \cite{atkey13abstraction} have already
presented a polymorphic type system for expressing geometric
invariance properties similar to the

FIXME: area of triangle type

By extending System F$\omega$, we get to

\paragraph{Contributions}

Our major contributions are threefold:
\begin{enumerate}
\item FIXME: reformulate AJK in terms of the reflexive graph model of
  relational parametricity in order to unify that paper with the
  natural extension of Reynolds' relational parametricity to indexed
  types.
\item FIXME: Present an extension of System F$\omega$ suitable for
  writing invariant Lagrangians. This requires special attention to
  ensure that all the functions that we can write are actually
  continuous and differentiable. Hence we include a special type of
  smooth, invariant functions between cartesian spaces.
\item FIXME: Present many examples of 
\end{enumerate}

\paragraph{Outline}

@@

FIXME: Define one-parameter groups

\paragraph{Invariance of the Action} Let
\begin{displaymath}
  \mathcal{S}[q;a;b] = \int_a^b L(t,q(t),\dot{q}(t)) \mathit{dt}
\end{displaymath}
be the action of some physical system described by the Lagrangian
$L$. Assume a differentiable invertible function $\tau : \mathbb{R}
\to \mathbb{R}$ that transforms time in some way and a function
$\sigma : \mathbb{R}^n \to \mathbb{R}^n$ that transforms the vector of
generalised coordinates $\vec{q}$ to another vector of generalised
coordinates $\sigma(\vec{q})$.

We say that the action $\mathcal{S}$ is invariant under the
transformations $\tau$ and $\sigma$ if it is the case that for all
$a$, $b$ and $q$ that:
\begin{equation}\label{eq:invariance}
    \int_a^b L(t,q(t),\dot{q}(t)) \mathit{dt}
    = \int_{\tau(a)}^{\tau(b)} L(s, q^*(s), \dot{q^*}(s)) \mathit{ds}
\end{equation}
where $q^*(s) = (\sigma \circ q \circ \tau^{-1})(s)$ is the path $q$
transformed by $\tau$ and $\sigma$.  FIXME: Remark on the use of $t$
and $s$ in the integrals. By change of variables, the right hand
integral in the above equation is equal to the following integral:
\begin{displaymath}
  \int_a^b L(\tau(t), q^*(\tau(t)), \dot{q^*}(\tau(t)))\cdot \dot{\tau}(t) \mathit{dt}
\end{displaymath}
Since the endpoints $a$ and $b$ are arbitrary

@@

The interpretation of kinds as reflexive graphs, and small kinds as
small reflexive graphs is a key property of our semantics that we will
maintain as we add additional kinds in
Sections~\ref{sec:discrete-kinds} and~\ref{sec:groupoid-kinds},
below. We state this as Property~\ref{property:semantic-kinds}:

@@

In Section~\ref{sec:refl-graphs-for-fomega}, we only had a single base
kind: the kind $*$ of proper types, with a specific intepretation as
the reflexive graph of sets and relations. We now describe two
families of base kinds with interpretations that are particular sorts
of reflexive graphs. In this section, we look at discrete kinds; kinds
whose reflexive graph interpretations are such that the reflexive
edges are the only edges between objects. In the following section
(Section~\ref{sec:groupoid-kinds}), we look at groupoid kinds, where
edges are composable and invertible.

Discrete kinds can be seen as the natural way of lifting base types up
to the kind level.

@@

\begin{mathpar}
  \inferrule*
  { }
  {\Theta \vdash \mathit{zero} : \mathsf{nat}}

  \inferrule*
  {\Theta \vdash A : \mathsf{nat}}
  {\Theta \vdash \mathit{succ}~A : \mathsf{nat}}

  \inferrule*
  {\Theta \vdash A : \mathsf{nat} \\
    \Theta \vdash B : \kappa \\
    \Theta \vdash C : \kappa \to \kappa}
  {\Theta \vdash \mathit{natrec}~A~B~C : \kappa}
\end{mathpar}

@@

\subsection{Groupoid Kinds}
\label{sec:groupoid-kinds}

@@

\paragraph{The additive group of integers}

\paragraph{Matrix groups}

GL and O, with group operations

$\mathit{ortho}_n : \mathsf{O}(n) \to \mathsf{GL}(n)$
$\mathit{scale}_n : \mathsf{GL}(1) \to \mathsf{GL}(n)$

\paragraph{Translation groups}

$\mathit{exp} : \mathsf{T}(1) \to \mathsf{GL}(1)$
$\underline{c}\cdot- : \mathsf{Z} \to \mathsf{T}(1)$

\paragraph{Groupoid of Cartesian Spaces}

\subsection{Why not Groupoids Everywhere?}

Explain why we aren't using groupoids as kinds: don't need to, doesn't
directly generalise relational parametricity, see Robinson's
paper. Distance-indexed types.

@@

%\subsection{Extending System F$\omega$ for Classical Mechanics}

@@

\subsection{Extending System F$\omega$ with Smooth Functions}

\begin{enumerate}
\item We have already done the matrix kinds and the CartSp kind back
  in the previous section. Add $n-ary$ products for CartSp.
  \begin{displaymath}
    \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
      C^\infty &:& \mathsf{CartSp} \to \mathsf{CartSp} \to * \\
      \mathbb{R}^n &:& \mathsf{GL}(n) \to \mathsf{T}(n) \to \mathsf{CartSp} \\
%      (\times) &:& \mathsf{CartSp} \to \mathsf{CartSp} \to \mathsf{CartSp} \\
      \mathit{vec} &:& \mathsf{nat} \to \mathsf{CartSp} \to \mathsf{CartSp}
    \end{array}
\end{displaymath}
\item Note that the $n$ in $\mathbb{R}^n$ is static
\item Add the type former for the smooth function space, and all the
  relevant constants
\item Give a (very) small example, and complain that it is very
  difficult to write anything in this point-free style.
\end{enumerate}

Administrative things: composition, products

\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    (+) &:&
    \begin{array}[t]{@{}l}
      \forall g \mathord: \mathsf{GL}(n), t_1,t_2 \mathord: \mathsf{T}(n). \\
      \hspace{1.5cm} C^\infty(\mathbb{R}^n\langle g, t_1 \rangle \times \mathbb{R}^n\langle g, t_2 \rangle, \mathbb{R}^n\langle g, t_1 + t_2 \rangle)\\
    \end{array} \\
    (-) &:&
    \begin{array}[t]{@{}l}
      \forall g \mathord: \mathsf{GL}(n), t_1,t_2 \mathord: \mathsf{T}(n). \\
      \hspace{1.5cm} C^\infty(\mathbb{R}^n\langle g, t_1 \rangle \times \mathbb{R}^n\langle g, t_2 \rangle, \mathbb{R}^n\langle g, t_1 - t_2 \rangle)\\
    \end{array} \\
    \sin &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, 2\pi \cdot z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
    \cos &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, 2\pi \cdot z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
    \mathrm{exp} &:& \forall t \mathord: \mathsf{T}(1).~C^\infty(\mathbb{R}\langle 1, t \rangle, \mathbb{R}\langle \mathit{exp}~t, 0 \rangle)
  \end{array}
\end{displaymath}

\subsection{A Surface Syntax for Smooth Functions}

\begin{enumerate}
\item Kind of cartesian spaces, parameterised by a matrix group and a
  translation group
\item Smooth functions between spaces: motivate this by looking at the
  requirements of the Euler-Lagrange equations.
\item Primitives for dealing with the geometry of the spaces (inner
  product, vector space operations, cross product(?), trigonometric
  functions (at least sin and cos)).
\end{enumerate}

\begin{displaymath}
  \kappa ::= \cdots \sepbar \mathsf{CartSp} \sepbar \mathsf{T}(n) \sepbar \mathsf{GL}(n) \sepbar \mathsf{O}(n) \sepbar \mathsf{Z}
\end{displaymath}

@@

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle \mathit{scale}_n(s)\mathit{ortho}_n(o), 0 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle \mathit{scale}_n(s)\mathit{ortho}_n(o), 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 \cdot e_2 : \mathbb{R}\langle\mathit{scale}_n(s)^n, 0 \rangle}
  \end{mathpar}
  \textbf{Transcendental functions}
  \begin{mathpar}
    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 1, t \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{exp}~e : \mathbb{R}\langle \mathit{exp}~t, 0 \rangle}

@@
